---
title: "GY7702 Assignment"
author: "Elliot Greatrix"
date: "04/11/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
```

The libraries that were used in this assignment: 
```{r load_libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(palmerpenguins)
library(dplyr)
library(knitr)
library(readr)
library(tinytex)
```


# Question 1

A vector of 25 numbers between 1 and 7 are listed below. These values represent
answers to a survey question 

## NA 3 4 4 5 2 4 NA 6 3 5 4 0 5 7 5 NA 5 2 4 NA 3 3 5 NA

*  1 = Completely disagree
*  2 = Disagree
*  3 = Somehow disagree
*  4 = Neither agree or disagree
*  5 = Somehow agree
*  6 = Agree
*  7 = Completely agree
* NA = missing value


## Question 1.1 

Write the code necessary to check whether all participants to the survey either 
completely disagree or completely agree, once the missing values are excluded

First, let's create the vector. 
```{r}
num <- c(NA, 3, 4, 4, 5, 2, 4, NA, 6, 3, 5, 4, 0, 5, 7, 5, NA, 5, 2, 4, NA, 3, 3, 5, NA)
```

Then, remove missing values.
```{r}
num <- na.omit(num)
```

And finally check whether all students either completely agree (= 7).
```{r}
students_agree_complete <- all(num == 7)
```

Or, completely disagree (=1).
```{r}
students_disagree_complete <- all(num == 1)
```


## Question 1.2

Write the code necessary to extract the indexes related to the participants 
in the survey who at least somehow agree or more.

In the vector *num*, *which* students selected values between 5 and 7. 
```{r}
students_agree <- which(num %in% c(5:7))
```

# Question 2

## Question 2.2

The library *palmerpenguins* is a data set of penguin characteristics in the table *penguins*. This data was made available by Dr. Kristen Gorman.

The library is already installed and loaded, so let's dive into some analysis.   
 
Some of the penguins in table hold NA values for certain attributes. Due to the analysis that we are going to do, these need to be removed
```{r}
pen <- na.omit(penguins)
```

Ultimately, We are going to write the code necessary to create a table showing species, island, bill length and body mass of the 10 Gentoo penguins in the penguins table with the highest body mass. 

As this task requires multiple steps, it will be more efficient to start a 'pipe'.
Lets send the 'na omitted' table, *pen* down the pipe.
```{r}
pen %>%
```


First, let's select all the attributes that the questions asks for
```{r}
 select(species, island, bill_length_mm, bill_depth_mm,)%>%
```

and select all the 'Gentoo' penguins.
```{r}
filter(species == "Gentoo")%>%
```

The question asks to return the 10 Gentoo penguins with the highest body mass. 
This can be achieved with *slice_max*
```{r}
slice_max(body_mass_g, n = 10)
```

## Question 2.3

Next, we're going to write the code necessary to create a table showing the 
average bill length per island, ordered by average bill length.

Similarly, lets open a new pipe operator
```{r}
pen %>%
```

and select the appropriate attributes.
```{r}
select(species, island, bill_length_mm, bill_depth_mm,)%>%
```

Order by *bill_length_mm*. By default *arrange* sorts in ascending order.
```{r}
arrange(bill_length_mm)%>%
```

As the question asks for *per island*, we can use *group_by* to execute the 
average for each island
```{r}
group_by(island)%>%
```

```{r, echo=FALSE}
rm(list = ls())
```





