---
title: "GY7702 Assignment"
author: "Elliot Greatrix"
date: "04/11/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
```

```{r load_libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(palmerpenguins)
library(dplyr)
library(knitr)
library(readr)
library(tinytex)
```


# Question 1

A vector of 25 numbers between 1 and 7 are listed below. These values represent
answers to a survey question 

## NA 3 4 4 5 2 4 NA 6 3 5 4 0 5 7 5 NA 5 2 4 NA 3 3 5 NA

*  1 = Completely disagree
*  2 = Disagree
*  3 = Somehow disagree
*  4 = Neither agree or disagree
*  5 = Somehow agree
*  6 = Agree
*  7 = Completely agree
* NA = missing value


## Question 1.1 

Write the code necessary to check whether all participants to the survey either 
completely disagree or completely agree, once the missing values are excluded

First, let's create the vector. 
```{r}
num <- c(NA, 3, 4, 4, 5, 2, 4, NA, 6, 3, 5, 4, 0, 5, 7, 5, NA, 5, 2, 4, NA, 3, 3, 5, NA)
```

Then, remove missing values.
```{r}
num <- na.omit(num)
```

And finally check whether all students either completely agree (= 7).
```{r}
students_agree_complete <- all(num == 7)
```

Or, completely disagree (=1).
```{r}
students_disagree_complete <- all(num == 1)
```


## Question 1.2

Write the code necessary to extract the indexes related to the participants 
in the survey who at least somehow agree or more.

In the vector *num*, *which* students selected values between 5 and 7. 
```{r}
students_agree <- which(num %in% c(5:7))
```


Clear the environment 
```{r}
rm(list = ls())
```





